---
title: "Tidal Creek Assessment"
csl: stylefile.csl
bibliography: tidalrefs.bib
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Tidal Creek Assessment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = F, warning = F, 
  fig.align = 'center', 
  cache = F
)

# libraries
library(tbeptools) 
library(bookdown)
library(dplyr)
library(knitr)
library(mapview)
library(plotly)
library(haven)
library(sf)

sf::st_crs(tidalcreeks) <- 4326
# spelling::spell_check_files("vignettes/tidalcreeks.Rmd")
```


```{r}

setwd("E:/To Marcus/Feed dashboard")

#iwrraw<-read_sas("iwrraw_creeks61.sas7bdat")
#creek_pop<-read_sas("creek_pop61.sas7bdat")

#save(iwrraw,file="iwrraw.rdata")
#save(creek_pop,file="creek_pop.rdata")

Creek_jei<- sf::read_sf(dsn = ".", layer = "TidalCreek_All_Line") 

```

## The Dashboard functions run using code below after loading iwrraw into the working environment but im not sure that this isnt going back to the old file? Not sure where iwrraw is stored in the main repo. 

never the less, these answers wont be the same because they use the old creek pop file and the  old line coverage where in some cases the line file was improved and creek length may have changed. 

```{r}
results <- anlz_tdlcrk(tidalcreeks, iwrraw)
results
```


```{r, eval = T, cache = F}
show_tdlcrk(results)
```

