---
title: "Seagrass Transect Data"
csl: stylefile.csl
# bibliography: seagrasstransect.bib
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Seagrass Transect Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = F, warning = F, 
  fig.align = 'center'
)

library(tbeptools)
library(mapview)

# bib_scrp(here('vignettes', 'seagrasstransect.Rmd'), bib_new = here('vignettes', 'seagrasstransect.bib'), ext_bib = 'https://raw.githubusercontent.com/tbep-tech/tbep-refs/master/bibs.bib')

# spelling::spell_check_files(here::here("vignettes", "seagrasstransect.Rmd"))
```

Each year, TBEP partners collect seagrass transect data at fixed locations in Tampa Bay.  Data have been collected since the mid 1990s and are hosted online at the [Tampa Bay Water Atlas](https://www.tampabay.wateratlas.usf.edu/) by the University of South Florida Water Institute.  A limited set of functions are available in tbeptools for downloading and analyzing these data. 

## Transect locations

There are two datasets included in tbeptools that show the transect locations that are actively monitored in Tampa Bay.  The `trnpts` dataset is a point object for the starting location of each transect and the `trnlns` dataset is a line object showing the approximate direction and length of each transect beginning at each point in `trnpts`.  Each dataset also includes the `MonAgency` column that indicates which monitoring agency collects the data at each transect. 

```{r}
trnpts
trnlns
```

The two datasets are `sf()` (simple features) objects and are easily mapped with `mapview()` to view their locations.

```{r}
mapview(trnpts, label = trnpts$lab, zcol = 'MonAgency', lwd = 0, legend = F, homebutton = F) + 
  mapview(trnlns, zcol = 'MonAgency', homebutton = F, layer.name = 'Monitoring Agency', lwd = 4)
```

## Accessing the transect data

The transect data can be downloaded from the Water Atlas using the `read_transect()` function.  The only required argument for this function is `training`, which indicates if you want to download training data or the complete dataset, i.e., `training = TRUE` or `training = FALSE` (default).  In the former case, a small dataset is downloaded that includes only data collected during an annual training event. These are used primarily to assess precision among different training crews. The data are downloaded as a JSON object and formatted internally using the `read_formtransect()` function. 

```{r}
# import JSON
trndat <- read_transect(training = TRUE)

# view the data
trndat
```

The `show_compplot()` function is used to compare training data between crews for a selected species (`species` argument) and variable (`varplo` argument). 

```{r, fig.height = 5, fig.width = 7}
# plot
show_compplot(trndat, site = '1', species = 'Halodule', varplo = 'Abundance')
```

Change the `training` argument to `FALSE` to download the entire transect database.  This may take a few seconds.

```{r}
# import entire transct dataset as JSON
trndat <- read_transect(training = FALSE)

# view the data
trndat
```

The columns in the transect database describe the crew (`Crew`), the monitoring agency (`MonitoringAgency`), sample date (`Date`), transect name (`Transect`), the meter location along the transect (`Site`, m), depth at the site (`Depth`, cm), Seagrass species (`Savspecies`), distance of the seagrass edge on the transect (`SeagrassEdge`, m), the seagrass variable (`var`), average value of the variable (`aveval`), and standard deviation of the variable if appropriate (`sdval`).  
