---
title: "Tampa Bay Benthic Index"
csl: stylefile.csl
bibliography: tbbirefs.bib
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Tampa Bay Benthic Index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = F, warning = F, 
  fig.align = 'center'
)

# libraries
library(tbeptools)
library(bookdown)
library(patchwork)
library(dplyr)
library(knitr)
library(mapview)
library(ggplot2)
library(sf)
library(tibble)

# spelling::spell_check_files("vignettes/tbbi.Rmd")
```

## Background

The Tampa Bay Benthic Index (TBBI) [@tbep0620;@tbep0106]) is an assessment method that quantifies the ecological health of the benthic community in Tampa Bay.  The index provides a complementary approach to evaluating environmental condition that is supported by other assessment methods currently available for the region (e.g., water quality report card, nekton index, etc.). The tbeptools package includes several functions described below to import data required for the index and plot the results to view trends over time.  Each of the functions are described in detail below. 

The TBBI uses data from the Tampa Bay Benthic Monitoring Program as part of the Hillsborough Country Environmental Protection Commission (EPC).  The data are updated annually on a FTP site maintained by EPC, typically in December after Summer/Fall sampling.  This is the same website that hosts water quality data used for the [water quality report card](https://tbep-tech.github.io/tbeptools/articles/intro.html).  The required data for the TBBI are more extensive than the water quality report card and the data are made available as a zipped Microsoft Access Database.  The process for downloading and working with the data are similar as for the other functions in tbeptools, except a few additional steps to connect and extract tables from the Access database are required.  Because of this, __the data import functions have only been tested on Windows using 64-bit R.__ You will need to download and install a [Microsoft Access driver](https://www.microsoft.com/en-us/download/details.aspx?id=54920) to import the data.

## Data import and included datasets

After installing a [Microsoft Access driver](https://www.microsoft.com/en-us/download/details.aspx?id=54920), data for calculating TBBI scores can be imported into the current R session using the `read_importbenthic()` function.  This function downloads the latest Access database from the EPC FTP site if the data have not already been downloaded to the location specified by the input arguments.  

The Access database is hosted on the FTP site as a zipped file.  First create a character path for the location where you want to download the zipped file. If one does not exist, specify a location and name for the downloaded file. Make sure you do not use the '~' path expansion because this will confuse the database connection in a later step.  Here, we want to put the file `benthic.zip` at the location `C:/path/to/` (a made up place for this example).

```{r, eval = F}
path <- 'C:/path/to/benthic.zip'
benthicdata <- read_importbenthos(path)
```

Running the above code will return the following error: 

```{r, eval = F}
#> Error in read_importbenthic() : File at path does not exist, use download_test = TRUE
```

We get an error message from the function indicating that the file is not found. This makes sense because the file does not exist yet, so we need to tell the function to download the latest file. This is done by changing the `download_latest` argument to `TRUE` (the default is `FALSE`).

```{r, eval = F}
benthicdata <- read_importbenthic(path, download_latest = T)
```
```{r, eval = F}
#> File C:/path/to/benthic.zip does not exist, replacing with downloaded file...

#> trying URL 'ftp://ftp.epchc.org/EPC_ERM_FTP/Benthic_Monitoring/EPC DataSubmittals.zip' length 37122877 bytes (35.4 MB)
```

Now we get an indication that the file on the server is being downloaded. When the download is complete, we’ll have the data downloaded and saved to the `benthicdata` object in the current R session.

If we try to run the function again after downloading the data, we get the following message. This check is done to make sure that the data are not unnecessarily downloaded if the current file matches the file on the server.

```{r, eval = F}
benthicdata <- read_importbenthic(path, download_latest = T)
```
```{r, eval = F}
#> File is current..
```

Every time that tbeptools is used to work with the benthic data, `read_importbenthic()` should be used to import the data. You will always receive the message `File is current...` if your local file matches the one on the server. However, data are periodically updated and posted on the server. If `download_latest = TRUE` and your local file is out of date, you will receive the following message:

```{r, eval = F}
#> Replacing local file with current...
```

Because the downloaded data are a zipped file of an Access database, a few extra steps are needed to process the data.  These are all handled automatically by the functions, but it does affect how you might use the data import function. The function automatically downloads the zipped file, unzips it, and establishes a connection to the Access database.  If you have already downloaded the zip file and extracted the .mdb file, you can connect to the latter directly by changing the \code{path} argument with the appropriate file extension. 

```{r, eval = F}
benthicdata <- read_importbenthic('C:/path/to/benthic.mdb')
```

The Access database also is quite large and you can use a temporary file path for the download and unzipping location.  You can use the `tempfile()` function with the .zip file extension for the `path` argument. Because a temporary file is used, you must always include `download_latest = TRUE`, i.e., there is no previously existing file. 

```{r, eval = F}
benthicdata <- read_importbenthic(tempfile(fileext = '.zip'), download_latest = T)
```

## Calculating TBBI scores

After the data are imported, you can view them from the assigned object.  The data are provided as a nested tibble that includes three different datasets: station information, field sample data (salinity), and detailed taxa information.  

```{r}
benthicdata
```

The individual datasets can be viewed by extracting them from the parent object using the `deframe()` function from the tibble package. 

```{r}
# see all
deframe(benthicdata)

# get only station dat
deframe(benthicdata)[['stations']]
```

The `anlz_tbbiscr()` function uses the nested `benthicdata` to estimate the TBBI scores at each site. The TBBI scores typically range from 0 to 100 and are grouped into categories that describe the general condition of the benthic community.  Scores less than 73 are considered "degraded", scores between 73 and 87 are "intermediate", and scores greater than 87 are "healthy".  Locations that were sampled but no organisms were found are assigned a score of zero and a category of "empty sample".  The total abundance (`TotalAbundance`, organisms/m2), species richness (`SpeciesRichness`) and bottom salinity (`Salinity`, psu) are also provided.  Some metrics for the TBBI are corrected for salinity and bottom measurements taken at the time of sampling are required for accurate calculation of the TBBI.

```{r}
tbbiscr <- anlz_tbbiscr(benthicdata)
tbbiscr
```

## Plotting results

The TBBI scores can be viewed as annual averages for each bay segment using the `show_tbbimatrix()` function.  The `show_tbbimatrix()` plots the annual bay segment averages as categorical values in a conventional "stoplight" graphic. A baywide estimate is also returned, one based on all samples across all locations ("All") and another weighted by the relative surface areas of each bay segment ("All (wt)"). The input to `show_tbbimatrix()` function is the output from the `anlz_tbbiscr()` function.  

```{r, fig.height = 7, fig.width = 3}
show_tbbimatrix(tbbiscr)
```

The matrix can also be produced as a [plotly](https://plotly.com/r/) interactive plot by setting `plotly = TRUE` inside the function. 

```{r, fig.height = 7, fig.width = 4}
show_tbbimatrix(tbbiscr, plotly = T)
```

# References
